### 2025-03-01 paper/main.tex
Hainje formula; to-do items for us.

% Note to self: Please don't work on this, ever.

% to-do items
% -----------
% - Show that there is NOT an exact solution at polynomial degree 2 order.
% - Make the code output something that could be executed by Mathematica. Then prove using Mathematica that the formula reduces to Heron's formula.
% - Make the code output explicit expressions for phi and rho, and include them in section 3.
% - Needs a clear tikz figure showing the corners and notation.
% - What are the 3-dimensional and d-dimensional generalizations of this formula? Are they hard to find? YES I BET THEY ARE.
% - Discuss alternatives like Novara's; And alternatives like Kaze's.

@@

% typesetting issues
\newcommand{\documentname}{\textsl{Note}}
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\addtolength{\topmargin}{-0.20in}
\addtolength{\textheight}{1.40in}
\sloppy\sloppypar\raggedbottom\frenchspacing
\pagestyle{myheadings}
\markboth{foo}{Hogg \& Villar / A new formula for the area of a triangle}

@@

\section*{\raggedright
A new formula for the area of a triangle:
Useless, but explicitly in Deep Sets and map--reduce form}

\noindent \textbf{David W. Hogg}\footnote{The authors thank Alexander Novara (NYU), Ben Wandelt (Sorbonne), and Kaze Wong (Flatiron) for valuable discussions.}\footnote{Email at \texttt{<david.hogg@nyu.edu>}. DWH is at the Center for Cosmology and Particle Physics, Department of Physics, New York University; the Max-Planck-Insitut f\"ur Astronomie; and the Center for Computational Astrophysics, Flatiron Institute. The Flatiron Institute is a division of the Simons Foundation.} and
\textbf{Soledad Villar}\footnote{SV is at the Department of Applied Mathematics and Statistics, Johns Hopkins University; and the Mathematical Institute for Data Science, Johns Hopkins University.}

\paragraph{Abstract:}
It is known that any permutation-invariant function of data points $r_i$
can be written in the form $\rho(\sum_i\phi(r_i))$,
where $\rho$ and $\phi$ are (probably nonlinear) functions, and the function $\phi$ might output a high-dimensional latent encoding of each point $r_i$.
This form---known in the machine-learning literature as Deep Sets---is a map--reduce form, because the $\phi$ executions can be performed asynchronously and the sums can be performed hierarchically or in any order.
The area of a triangle is a permutation-invariant function of the locations $r_i$ of the three corners $1\leq i\leq 3$.
We find, possibly for the first time, the formula for the area of a triangle, given the positions of the three corners, that is explicitly in Deep Sets form.
It is the square-root of a permutation-invariant quartic polynomial.
No insights of any kind were gained by this work.

@@

The theory of permutation-invariant functions says (CITE) that any per\-mu\-tation-invariant function (under conditions??) of $N$ objects $r_i$ (with $1\leq i\leq N$) can be approximated (to arbitrary precision??) with an expression of the form
\begin{equation}
    \rho(\sum_{i=1}^N\phi(r_i)) \label{eq:DeepSets}
\end{equation}
where $\rho$ and $\phi$ are (possibly very nonlinear) functions,
and the $r_i$ are the data points that can be permuted without changing the result.
This result---which we associate with the subject \emph{Deep Sets} (CITE)---is not at all obvious:
For example, even though Heron's formula \eqref{eq:Heron} is permutation-invariant, it is not actually in the form \eqref{eq:DeepSets}, if the triangle is seen as being defined by its corners (or even its sides).

@@

Although the contribution of this paper is minimal, or even zero, it was inspired by real problems in physics.
For one very specific example, all known formulae for the gravitational potential energy of a system of $N$ point masses involve sums over pairs of points; they have not ever been written in the Deep Sets form \eqref{eq:DeepSets}.
In cosmology, all known estimators for $k$-point functions (correlation functions, bispectra, and so on) involve sums over $k$-tuples; once again they have not been written in the Deep Sets form.
Importantly, these kinds of problems cannot currently be performed (not even approximately) in any map--reduce framework; indeed they are very time-consuming because they involve sums over all elements of $N$-choose-$k$, which scales (badly) as $N^k$.
Because 3-point functions involve the shapes and areas of triangles, consideration of the 3-point function led to the question answered in this \documentname:
\emph{What is the Deep Sets (or map--reduce) expression for the area of a triangle?}
We deliver in \secref{sec:result}.

@@

HOGG: How to construct all the invariant polynomials.

@@

@@

Why not simplify the prefactors?

What's the generalization to $d$ dimensions?

Where does the factor of 23 come from?

What happens if we try to do $A$ as a quadratic instead of $A^2$ as a quartic?

What happens if you hold the formula fixed, but use more than $N=3$ points?
The formula still returns a squared area.
That's the area of what?

>>>>

% Copyright 2025 the authors.

% to-do items
% -----------
% - CH: Include the d=3, d=4 solutions and comments on d>4.
% - CH: Rewrite to remove the Hogg solution.
% - CH: Show that there is NOT an exact solution at polynomial degree 2 order.
% - CH: Needs a clear tikz figure showing the corners and notation. Like an infographic.
% - HOGG: Discuss alternatives like Novara's; And alternatives like Kaze's. Positional encodings! CH: Describe the connection to positional encodings (in, eg, transformers). Oh and cosmology and power spectrum and bispectrum.

@@

% typesetting issues
\newcommand{\documentname}{\textsl{Note}}
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\setlength{\textwidth}{5.50in}
\setlength{\oddsidemargin}{3.25in}\addtolength{\oddsidemargin}{-0.5\textwidth}
\addtolength{\topmargin}{-0.20in}
\addtolength{\textheight}{1.40in}
\sloppy\sloppypar\raggedbottom\frenchspacing
\pagestyle{myheadings}
\markboth{foo}{\sffamily Hainje, Hogg, \& Villar / A new formula for the area of a triangle}

\usepackage{natbib}
\usepackage{aas_macros}
\bibliographystyle{aasjournal}
\setcitestyle{aysep={}}
\usepackage{setspace}
\setlength{\bibsep}{0pt plus 0.3ex}

@@

\section*{\raggedright
A new formula for the area of a triangle:
Useless, but explicitly in Deep Sets and map--reduce form\footnote{%
The authors thank Alexander Novara (NYU), Ben Wandelt (Sorbonne), and Kaze Wong (Flatiron) for valuable discussions.}}

\noindent \textbf{Connor Hainje},\footnote{Email at \href{mailto:connor.hainje@nyu.edu}{connor.hainje@nyu.edu}. CH is at the Center for Cosmology and Particle Physics, Department of Physics, New York University.}
\noindent \textbf{David W. Hogg},\footnote{DWH is at the Center for Cosmology and Particle Physics, Department of Physics, New York University; the Max-Planck-Insitut f\"ur Astronomie; and the Center for Computational Astrophysics, Flatiron Institute. The Flatiron Institute is a division of the Simons Foundation.} and
\textbf{Soledad Villar}\footnote{SV is at the Department of Applied Mathematics and Statistics, Johns Hopkins University; and the Mathematical Institute for Data Science, Johns Hopkins University.}

\paragraph{Abstract:}
It is known that any permutation-invariant function of data points $r_i$
can be written in the form $\rho(\sum_i\phi(r_i))$,
where $\rho$ and $\phi$ are (probably nonlinear) functions, and the function $\phi$ might output a high-dimensional latent encoding of each point $r_i$.
This form---known in the machine-learning literature as Deep Sets---is a map--reduce form, because the $\phi$ executions can be performed asynchronously and the sums can be performed hierarchically or in any order.
The area of a triangle is a permutation-invariant function of the locations $r_i$ of the three corners $1\leq i\leq 3$.
We find, possibly for the first time, the formula for the area of a triangle, given the positions of the three corners, that is explicitly in Deep Sets form.
% It is the square-root of a permutation-invariant quartic polynomial involving the integer 368.
No insights of any kind were gained from this work.

@@

The theory of permutation-invariant functions says that any per\-mu\-tation-invariant function (under conditions??) of $N$ objects $r_i$ (with $1\leq i\leq N$) can be approximated (to arbitrary precision??) with an expression of the form \citep{zaheer+17-Deep-Sets}
\begin{equation}
    \rho(\sum_{i=1}^N\phi(r_i)) \label{eq:DeepSets}
\end{equation}
where $\rho$ and $\phi$ are (possibly very nonlinear) functions,
and the $r_i$ are the data points that can be permuted without changing the result.
This result---which we associate with the subject \emph{Deep Sets} (CITE)---is not at all obvious:
For example, even though Heron's formula \eqref{eq:Heron} is permutation-invariant, it is not actually in the form \eqref{eq:DeepSets}, if the triangle is seen as being defined by its corners (or even its sides).

@@

Although the contribution of this paper is minimal, or even zero, it was inspired by real problems in physics.
For one very specific example, all known exact formulae for the gravitational potential energy of a system of $N$ point masses involve sums over pairs of points (and also require taking differences of points); they have not ever been written in Deep Sets form \eqref{eq:DeepSets} to our knowledge.
CH: THAT SAID, THERE ARE MULTIPOLE-LIKE EXPRESSIONS THAT MIGHT BE IN DEEPSETS AND APPROXIMATE THIS CALCULATION; SEE THE DISCUSSION?
In cosmology, standard estimators for $k$-point functions (correlation functions, bispectra, and so on) involve sums over $k$-tuples; once again they have not been written in the Deep Sets form.
CH: DO THE PHILCOX PAPERS ADDRESS THIS PREVIOUS POINT?
Importantly, when these statistics are computed in their naive forms, they are very time-consuming because they involve sums over all elements of $N$-choose-$k$, which scales (badly) as $N^k$.
Because 3-point functions involve the shapes and areas of triangles, consideration of the 3-point function led to the question answered in this \documentname:
\emph{What is the Deep Sets (or map--reduce) expression for the area of a triangle?}
We deliver in \secref{sec:result}.

@@

In order to generate all the possible invariant monomials, we first note that they are made of products of factors of the form $F_{m,n} = \sum_i x_i^m y_i^n$.
The invariant monomials of degree $d$ are then all products $\prod_i F_{m_i, n_i}$ such that $\sum_i m_i + n_i = d$.

% HOGG: How to construct all the invariant polynomials.

% HAINJE: Given data of form $(x, y)$, invariant factors are $F_{m,n} = \sum_i x_i^m y_i^n$, and the invariant monomials are any products of these factors $\prod_{i} F_{m_i,n_i}$. Do we want/need to be more specific than that?

@@

HAINJE: Here it is:
\begin{gather}
    \phi(r_i) = (x_i, y_i, x_i\,y_i, x_i^2, y_i^2, x_i\,y_i^2, x_i^2\,y_i, x_i^2\,y_i^2) \\
    \begin{aligned}
    \rho(\textstyle\sum_{i=1}^{3}\phi(r_i)\equiv \vec{u}) = &
    - \tfrac{54}{368} u_8
    - \tfrac{258}{368} u_3^2
    + \tfrac{285}{368} u_4\,u_5
    + \tfrac{36}{368} u_2\,u_7
    + \tfrac{36}{368} u_1\,u_6 \\ &
    - \tfrac{101}{368} u_2^2\,u_4
    + \tfrac{148}{368} u_1\,u_2\,u_3
    - \tfrac{101}{368} u_1^2\,u_5
    + \tfrac{9}{368} u_1^2\,u_2^2.
    \end{aligned}
\end{gather}

@@

HOGG SAY: CH: EXPAND THIS INTO AN OPEN QUESTIONS LIST:
\begin{itemize}
\item CH should answer this question: What's the generalization to $d$ dimensions? That is, what if the three points are each in d dimensions?
\item questionable: What happens if we try to do $A$ as a quadratic instead of $A^2$ as a quartic?
\item What happens if you hold the formula fixed, but use more than $N=3$ points?
The formula still returns a squared area.
That's the area of what?
\item What about areas of quadrangles or larger polygons?
\item What about volumes and hypervolumes of larger objects?
\end{itemize}

\newpage
By inspection, we have also found a simpler form of the same formula, which is given by
\begin{align}
    4 \, A^2 = & \
        2 \, (\sum_{i=1}^{3} x_i ) \, (\sum_{i=1}^{3} y_i ) \, (\sum_{i=1}^{3} x_i \, y_i )
        \nonumber \\ &
        + 3 \, (\sum_{i=1}^{3} x_i^2 ) \, (\sum_{i=1}^{3} y_i^2 )
        - 3 \, (\sum_{i=1}^{3} x_i \, y_i )^2.
        \nonumber \\ &
        - \, (\sum_{i=1}^{3} x_i )^2 \, (\sum_{i=1}^{3} y_i^2 )
        - \, (\sum_{i=1}^{3} y_i )^2 \, (\sum_{i=1}^{3} x_i^2 )
\end{align}
In explicit Deep Sets notation, this can be written
\begin{align}
    \phi(r_i) &= \begin{bmatrix}
        x_i & y_i & x_i\,y_i & x_i^2 & y_i^2 \end{bmatrix}^\top
        \equiv \vec{u}^\top, \\
    \rho(\sum_{i=1}^{3} \phi(r_i))
    &= \tfrac{2}{4} u_1\,u_2\,u_3
    + \tfrac{3}{4} u_4\,u_5
    - \tfrac{1}{4} u_1^2\,u_5
    - \tfrac{1}{4} u_2^2\,u_4
    - \tfrac{3}{4} u_3^2.
\end{align}

The two are related by
\begin{equation}
    4\,u_1\,u_6 + 4\,u_2\,u_7 - 6\,u_8 = 
    -u_1^2 u_2^2 + 4 u_1 u_2 u_3 - 2 u_3^2 + u_2^2 u_4 + u_1^2 u_5 - u_4 u_5 ;
\end{equation}
namely, plugging this into \eqref{eq:result} causes it to reduce immediately into this simpler form.

This form was not found by the symbolic regression of \secref{sec:method} because the regression makes use of \verb+lstsq+ with more parameters than data points. In this regime, the solution that minimizes the $\mathcal{L}_2$ norm of the weights (the coefficients of the invariant monomials) is returned. The $\mathcal{L}_2$ norm of the weights of our initial formula is $2127 / 1472 \approx 1.445$, whereas the $\mathcal{L}_2$ norm of the weights of this simpler formula is $3/2 = 1.5$.

HAINJE: Would lasso regression return my formula? Is there a yet simpler formula?

\newpage

Define the ``invariant factor'' in $d$ dimensions
\begin{equation}
    F_{m_1 \, m_2 \, \cdots} \equiv \sum_{1=1}^{3} \prod_{k=1}^{d} (\vec{r}_{i})_{k}^{m_k}.
\end{equation}
This is the sum over all three particles of the product of the $d$ components of each $\vec{r}_i$, each component raised to a power. For example, in 2D, $d=2$ and this expands to
\begin{equation}
    F_{\ell m} = \sum_{i=1}^{3} x_i^\ell \, y_i^m.
\end{equation}
In 3D, $d=3$ and this gives
\begin{equation}
    F_{\ell m n} = \sum_{i=1}^{3} x_i^\ell \, y_i^m \, z_i^n.
\end{equation}
Then, the area formula found above can be extended to more dimensions with the following formulae.

For reference, the 2D formula was
\begin{equation}
4 A_{\rm 2D}^2 = +3 F_{02} F_{20} -3 F_{11}^2 - F_{20} F_{01}^2 +2 F_{11} F_{10} F_{01} - F_{02} F_{10}^2.
\end{equation}
In 3D, we have
\begin{equation}
\begin{aligned}
4 A_{\rm 3D}^2 = &
+3 F_{020} F_{002} 
+3 F_{200} F_{002} 
+3 F_{020} F_{200} \\&
-3 F_{011}^2 
-3 F_{101}^2 
-3 F_{110}^2 \\&
+2 F_{011} F_{001} F_{010} 
+2 F_{100} F_{001} F_{101}
+2 F_{100} F_{010} F_{110} \\&
-F_{002} F_{010}^2 
-F_{002} F_{100}^2 
-F_{020} F_{001}^2 
-F_{020} F_{100}^2
-F_{200} F_{010}^2
-F_{200} F_{001}^2. 
\end{aligned}
\end{equation}
And in 4D, we have
\begin{equation}
\begin{aligned}
4 A_{\rm 4D}^2 = &
+3 F_{0002} F_{0020} 
+3 F_{0002} F_{0200} 
+3 F_{0002} F_{2000} 
\\&
+3 F_{0200} F_{2000} 
+3 F_{0200} F_{0020} 
+3 F_{0020} F_{2000} 
\\&
-3 F_{0011}^2
-3 F_{0101}^2
-3 F_{0110}^2
-3 F_{1001}^2 
-3 F_{1010}^2 
-3 F_{1100}^2
\\&
+2 F_{0010} F_{0011} F_{0001} 
+2 F_{0101} F_{0100} F_{0001} 
+2 F_{1001} F_{1000} F_{0001}
\\&
+2 F_{0010} F_{0100} F_{0110} 
+2 F_{0010} F_{1010} F_{1000}
+2 F_{1100} F_{0100} F_{1000} 
\\&
- F_{0020} F_{0001}^2
- F_{0200} F_{0001}^2
- F_{2000} F_{0001}^2
- F_{0002} F_{0010}^2
- F_{0200} F_{0010}^2
- F_{2000} F_{0010}^2
\\&
- F_{0002} F_{0100}^2 
- F_{0020} F_{0100}^2 
- F_{2000} F_{0100}^2
- F_{0002} F_{1000}^2 
- F_{0020} F_{1000}^2 
- F_{0200} F_{1000}^2
\end{aligned}
\end{equation}

\newpage
{\small\singlespacing\bibliography{main}}
